Lessons learned: creating tutorials to teach agent-based modelling to archaeologists

### 2025-06-06 paper.tex
update paper and correct some figures

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Lessons learned: creating tutorials teaching agent-based modelling for archaeologists},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Lessons learned: creating tutorials teaching agent-based modelling for archaeologists}
\author{}
\date{\vspace{-2.5em}March 17, 2025}

@@

While agent-based modelling importance in archaeological practice is steadily growing (Romanowska and Scherjon 2023) there are significant barriers to its wider adoption. One of the problems is the access to vocational training required to gain the necessary computer programming and modelling skills (Daems 2020). A survey of practitioners (Davies and Romanowska 2018) found significant deficiencies in the provision of ABM training; 76\% of modellers were primarily self-taught with some degree of secondary support, i.e.~27\% relied on peer support, only 7\% had any sort of formal training at a workshop or summer school, while only 14\% received training as part of a degree.

@@

The project sought to create Open (Open Access and Open Source) Educational Resources (OERs) for training in Agent-Based Modelling for Archaeologists. These OERs were to be HTML and JavaScript based so they could be used on any device with a browser, even if it is not connected to the internet. The full programme consists of five hands-on practical tutorials, taking learners through the process of creating the digital models. The modular architecture ensures that they can be independent or incorporated into other teaching tools like learning management systems. They are also deliberately `method agnostic' so they can be incorporated into any tools from self-teaching to MOOCs, and are supported by a set of support materials to encourage use of the OERs.

The activities the project implemented were:

\begin{itemize}
\item
  development of a hands-on vocational training programme, aligned with the recently published textbook \emph{Agent-Based Modelling for Archaeology: Simulating the Complexity of Societies} (Romanowska, Wren and Crabtree 2021);
\item
  conversion of the training programme into modular learning resources that are interactive, based on HTML and JavaScript standards;
\item
  creation of support materials to encourage use of the OERs;
\item
  creation of how-to guides to help trainers incorporate OERs into their teaching;
\item
  creation of a code and learning materials repository that will allow the project to continue sustainability after the European Union support has ended and to facilitate the creation of an open-source community to manage and improve the OERs in the future;
\item
  development of the framework for skills acquisition aligned with the Digital Skills Passport;
\item
  promotion through a series of multiplier events.
\end{itemize}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  a surge in professionals who can build, apply and critique archaeological simulation thus opening new research avenues for the discipline as a whole;
\item
  increased level of digital competence, specifically in advanced computer skills that are highly transferable to other industries. This includes technical skills such as programming but also ``soft'' skills related to computational thinking;
\item
  improved opportunities for archaeologists to work in other European countries (transnational mobility) through matching the skills being delivered with the European Commission's Digital Competence Framework for Citizens (2.2) and the sectoral Skills Passport for Archaeology, which means that it will be easier for individual workers to demonstrate their skills to employers across Europe;
\item
  increased opportunities for professional development by having access to professional training that can be incorporated into formal or non-formal/informal learning.
\end{enumerate}

For the secondary target audience of vocational training providers in archaeology e.g.~employers, colleges and not-for-profit organisations, the expected results and outcomes were:

@@

The ABMA learning materials assume that participants have learning skills at least at a secondary education level. They are suitable for participants with at least some background in archaeology, such as those typical among students after the first year of a Bachelor study in Archaeology, i.e., above EQF levels 4 or 5. In addition, the tutorials are aimed at professionals working in archaeology who have some experience with computer applications. We also assume that learners are proficient in English and have at least the Reading B1 level, but Reading B2 is recommended (Council of Europe 2020). The tutorials were also developed with a secondary aim to develop learners' general skills in the following competence areas: 1. Information and data literacy, 2. Communication and collaboration, 3. Digital content creation and, 5. Problem solving (European Commission, Joint Research Centre et al. 2022).

After initial debugging by the project team, the tutorials were tested on a wider sample of participants during several online and in-person events all of which were run by members of the project team (Table \ref{tab:events}). Recognising the difficulty of delivering practical training in an online environment (Microsoft Teams and Zoom) during the online events several members of the team were present which enabled participants to go to break-out rooms with an expert if they needed additional help or wanted to discuss things. Throughout the project we sought feedback from the participants and used it to improve the tutorials prior to the following events. The feedback was obtained through two surveys using Qualtrics (\url{https://www.qualtrics.com/}). The participants of the events were asked to respond to one survey before the start of the event and one after the end of the event. The first survey focused on the background of the participants and their original level of knowledge in relation to ABM (see appendix 1 for the questions). At the end of the event the participants were asked to answer a second survey. Questions of this survey were aimed at measuring the effectiveness of the tutorials and getting feedback on the workshop and tutorials (see appendix 2 for the questions). For some events participants had to register beforehand and we were able to send the pre-workshop survey by email. At other events no registration was possible or necessary and the pre-workshop survey was shared at the start of the event. The post-workshop survey was distributed using QR-codes or links at the end of the event.

@@

The surveys were analysed using R (R Core Team 2023). The following packages were used for analyses: ggplot2 (Wickham 2016), dplyr (Wickham et al. 2023), tidyr (Wickham, Vaughan and Girlich 2023), forcats(Wickham 2023a), lubridate (Grolemund and Wickham 2011) and stringr (Wickham 2023b). For the sake of reproducibility (Marwick 2017) this paper was written in R Markdown (\url{https://rmarkdown.rstudio.com/}), with data and code available at \url{https://github.com/ABMArchaeologists/ABMA_paper} (Visser et al. 2024b).

Over the course of the project three groups of students worked on the project during the Smart Solutions Semester at Saxion University of Applied Sciences. This is an interdisciplinary semester in which students of at least three different study-programmes or disciplines work together on a complex problem/project (\url{https://www.saxion.edu/business-and-research/collaborate-with-saxion/smart-solutions}). Within the semester students have to come up with new ideas or solutions to help a client with a wicked, or at least challenging, problem. The backgrounds of the various students we worked with were diverse: Applied Computer Science, Archaeology, Business Management Studies, Creative Business, Creative Media \& Game Technologies, and ICT. The aim of involving interdisciplinary groups in the project was to involve students in current research projects, to enable them to learn new things and share their innovative ideas and solutions with us. The students groups were supplementary to the deliverables of the project, but their new ideas were beneficial to the project. Their help during workshops was useful, but they also were engaged in developing educational material (Aalpoel et al. 2024), testing the tutorials, developing a style for the website (Jutte et al. 2024) and other materials.

@@

The ABMA tutorials are publicly hosted on GitHub (\url{https://github.com/ABMArchaeologists/ABMA_tutorials/}) and are citeable (Rocks-Macqueen et al. 2024). Anyone interested can download the tutorials and their source code. To increase the accessibility and visibility of the tutorials, they are also hosted on a static website (\url{https://abmarchaeologists.github.io/ABMA_website}) and can be followed without the need to download any software. The source code of the website is also openly available (Jutte et al. 2024).

Besides, the tutorial, the website provides: a link that can be used for feedback, basic explanations of agent-based modelling, an introductory presentation from one of the workshops, videos from an online course at Leiden University (Scherjon, Romanowska and Lambers 2019) and external resources that people can use to learn agent-based modelling. The website was mainly developed by a student group from Saxion University of Applied Sciences (Jutte et al. 2024).

@@

The final set of ABMA tutorials is available for download at \url{https://github.com/ABMArchaeologists/ABMA_tutorials} and is citeable (Rocks-Macqueen et al. 2024). The set consists of the following tutorials:

@@

In tutorial 2 the user will learn the basics of NetLogo by making their simulation on the Out of Africal dispersal of \emph{Homo sapiens} (Young and Bettinger 1995). The learner works through the basics of NetLogo syntax and learns how to set up a simulation and visualize its outcomes. This tutorial consists of nine lessons. This tutorial intends to guide the learner from the introduction to the beginner level of proficiency. The learning curve is relatively gentle. This is achieved by working with the NetLogo web interface that enables highlighting specific words and buttons and giving visual cues as to how to proceed. The tutorial introduces several commonly used built-in functions in Netlogo (called \emph{primitives}) and introduces the general structure of all models consisting of the initialization phase and the main simulation loop. It walks the participants through the most important features of the NetLogo world, such as dimensions, coordinates and the origin point of the grid on which the simulation runs. The learners gain their first stripes in programming by using simple loops and functions as well as conditional statements. The next level is achieved by introducing the topic of function definition and different types of variables. Exporting data produced by the model through plots is also explained.

In the third tutorial, the learner builds a simple trade model, again based on a published simulation (Romanowska 2018). Here learners expand their NetLogo programming skills with more complex syntactic structures as loops, lists and reporters. They are introduced to some key computational techniques like modular code development and debugging which become important alongside this increased coding complexity. This tutorial consists of seven lessons. It intends to guide the learner from the beginner to the intermediate level of proficiency. This is achieved by introducing standards of code development such as modular code, pseudocode, debugging and annotating the code. In addition, the tutorial introduces custom agent breeds, visualization with labels and reporters, plots and monitors. This takes the learner to a more advanced level, since it develops knowledge of the programming language combined with diagnosing unexpected behaviour and problem solving.

@@

A large proportion of the participants of the workshops gave us information using the survey before the workshops (172 of 368 participants). The respondents came from at least 40 countries that represented almost all continents (see Figure \ref{fig:nationality}). There was a clear skew towards participants from Europe. The number of female respondents slightly outnumbered the male ones and a small group did not share their gender, while two identified as non-binary (see Figure \ref{fig:gender-age}). The ages of the respondents ranged from below 20 to over 70, although the majority fell between 20 and 40 years old indicating that the audience was predominantly composed of early career archaeologists. It seems that the different events also had a slightly different distribution of both gender and age. For example, more older people attended the workshop at the CAA conference in April 2023 and more men were present during the workshop at the Reuvensdagen in November 2023. This might be due to the differences of audiences at the conferences. For the Reuvensdagen the number of participants of the workshop was relatively low and this variance might be due to chance.

@@

Most of the respondents reported having a level of computer literacy already including particular computational skills common in archaeology, for example working with word processors, GIS and spreadsheets (see Figure \ref{fig:computer-skills}), and 111 of the respondents had some prior knowledge of ABM, while 59 were completely new to the subject. The majority (158) had never applied ABM to their research before participating in the workshops with only 14 participants who had experience of developing archaeological ABM. It is interesting to note that many respondents (108) did know what kind of software was available for ABM (see Figure \ref{fig:abm-knowledge}).

@@

As shown above, the respondents had some knowledge on ABM in general, but did not know how to apply it or had never applied it before. The respondents were also asked how they rated the available resources on ABM (Figure \ref{fig:available-theory}). While 64 of them had no opinion on the subject, a proportion of the respondents (28) answered that they rated the available theory on ABM as limited and only a small group (39) as sufficient or better. This clearly shows the need for more and better educational material.

@@

\hypertarget{during-the-workshop}{%
\subsubsection{During the workshop}\label{during-the-workshop}}

During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials (Figure \ref{fig:workshop-impression}). Some participants engaged in discussion with the teachers to learn how they could apply ABM in their research of discuss possibilities for the application of ABM in general. Individual progress varied depending on skills and preferences of the participants. While bugs were still a problem in earlier workshops, participant-teacher interaction was more focused on content-related learning problems in later installments.

@@

A similar proportion of the participants of the workshops gave us information after the workshops (171 of 368 participants).

@@

The majority of the respondents (153) were enthusiastic about the teaching materials, with the majority rating them as excellent or good. The teachers were rated even better than the teaching materials, with more respondents (165) giving them the \emph{excellent} score (see Figure \ref{fig:rating-teaching}).

@@

The respondents generally liked the interactive step-by-step way of going through the tutorials, but also valued the interaction with the teachers strongly. The easy and intuitive introduction in ABM and NetLogo were also mentioned quite often. This was also observed during the workshops as most participants were working at their own pace and did not need much help. From a didactic point of view, the self-guided and self-paced way of going through the tutorials is very efficient for teachers and limits the frustration of students that may otherwise go too fast or fall behind the class when the tempo is dictated by the teacher.

@@

Beyond the creation of the OERs the project also worked on aligning the learning outcomes of the OERs with the Digital Competence Framework for Citizens and the Digital Archaeological Skills Passport.

@@

As mentioned above, the following competence areas were identified as relevant for the tutorials: 1. Information and data literacy, 2. Communication and collaboration, 3. Digital content creation and, 5. Problem solving (European Commission, Joint Research Centre et al. 2022). In these competence areas various competences are developed towards an advanced or even specialised level. A more extensive description of the Digital Competence Framework for Citizens in relation to Agent-Based Modelling for archaeologists is available elsewhere (Visser et al. 2024a).

For competence area 1. Information and data literacy: the tutorials mainly address 1.2 Evaluating Data, Information And Digital Content and 1.3 Managing Data, Information And Digital Content. In the ABMA educational materials, users learn about working with data, being critical of their digital content and how to model and manage the data in the context of an ABM. For tutorial 2 the user is expected to be able to find information using the NetLogo(Web) dictionary (\url{http://ccl.northwestern.edu/netlogo/bind/} or \url{http://ccl.northwestern.edu/netlogo/docs/dictionary.html}).

For Competence Area 2. Communication and collaboration the tutorials mainly address 2.1 Interacting Through Digital Technologies, because the users are constantly interacting with digital technologies, but communication is not relevant for the tutorials. However, during the workshops given in the course of the project, participants interacted with each other and the teachers.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Lessons learned: creating tutorials to teach agent-based modelling to archaeologists},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Lessons learned: creating tutorials to teach agent-based modelling to archaeologists}
\author{}
\date{\vspace{-2.5em}June 06, 2025}

@@

Although the importance of agent-based modelling in archaeological practice is steadily growing (Romanowska and Scherjon 2023) there are significant barriers to its wider adoption. While education on ICT skills related to techniques like GIS and databases have found their way into many study programmes (Garstki 2022; Visser, Zijverden and Alders 2016), the access to vocational training required to gain the necessary computer programming and modelling skills for ABM is limited. A survey of practitioners (Davies and Romanowska 2018) found significant deficiencies in the provision of ABM training; 70.8\% of modellers were primarily self-taught with some degree of secondary support, either by peers or academic supervisors. Only 13.8\% received training as a formal part of a degree. While ABM is taught outside of archaeology (e.g. Augustijn et al. 2019) and various handbooks exist in different domains (Acerbi, Mesoudi and Smolla 2023; Crooks et al. 2019; Hamill and Gilbert 2016; Railsback and Grimm 2019; Smaldino 2023; Wilensky and Rand 2015), it is lacking in most archaeological curricula due to for example lack of expertise in a university or competition with other subjects to be taught in a given curriculum. There are several online courses available, (e.g.~\url{https://acerbialberto.com/IBM-cultevo/}, \url{https://smaldino.com/wp/abm-courses/}, \url{https://forum.comses.net/t/educational-resources/9159}), but none are aimed at archaeology. The lack of training material on ABM for archaeologists, leaves this important method in digital archaeology at the ``fringes of the archaeological discipline'' (Daems 2020: 90) rather than moving it to the front lines.

@@

The ABMA project sought to create Open (Open Access and Open Source) Educational Resources (OERs) for training in Agent-Based Modelling for Archaeologists. These OERs were to be HTML- and JavaScript-based so they could be used on any device with a browser, even if it is not connected to the internet. The full programme consists of five hands-on practical tutorials, taking learners through the process of creating the digital models. The modular architecture ensures that they can be independent or incorporated into other teaching tools like learning management systems. They are also deliberately `method agnostic' so they can be incorporated into any tools from self-teaching to MOOCs, and are supported by a set of materials to encourage use of the OERs.

The activities that the project implemented included:

\begin{itemize}
\item
  development of a hands-on vocational training programme, aligned with the recently published textbook \emph{Agent-Based Modelling for Archaeology: Simulating the Complexity of Societies} (Romanowska, Wren and Crabtree 2021);
\item
  conversion of the training programme into modular learning resources that are interactive, based on HTML and JavaScript standards;
\item
  creation of support materials to encourage use of the OERs;
\item
  creation of how-to guides to help trainers incorporate OERs into their teaching;
\item
  creation of a code and learning materials repository that will allow the project to ensure sustainability after the European Union support has ended and to facilitate the creation of an open-source community to manage and improve the OERs in the future;
\item
  development of a framework for skills acquisition aligned with the Digital Skills Passport;
\item
  promotion through a series of multiplier events.
\end{itemize}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  a surge in professionals who can build, apply and critique archaeological simulation, thus opening new research avenues for the discipline as a whole;
\item
  increased level of digital competence, specifically in advanced computer skills that are highly transferable to other industries. This includes technical skills such as programming but also ``soft'' skills related to computational thinking;
\item
  improved opportunities for archaeologists to work in other European countries (transnational mobility) through matching the skills being delivered with the European Commission's Digital Competence Framework for Citizens (2.2) and the sectoral Skills Passport for Archaeology, which means that it will be easier for individual workers to demonstrate their skills to employers across Europe;
\item
  increased opportunities for professional development by having access to professional training that can be incorporated into formal or non-formal/informal learning.
\end{enumerate}

For the secondary target audience of vocational training providers in archaeology, e.g.~employers, colleges and not-for-profit organisations, the expected results and outcomes were:

@@

The ABMA learning materials assume that participants have learning skills at least at a secondary education level. Also, they are suitable for participants with at least some background in archaeology, such as those typical among students after the first year of a Bachelor study in Archaeology, i.e., above EQF levels 4 or 5. In addition, the tutorials are aimed at professionals working in archaeology who have some experience with computer applications. We also assume that learners are proficient in English and have at least the Reading B1 level, but Reading B2 is recommended (Council of Europe 2020). The tutorials were also developed with a secondary aim to develop learners' general skills in the following competence areas: 1. Information and data literacy, 2. Communication and collaboration, 3. Digital content creation and, 5. Problem solving (European Commission, Joint Research Centre et al. 2022).

After initial debugging by the project team, the tutorials were tested on a wider sample of participants during several online and in-person events, all of which were run by members of the project team (Table \ref{tab:events}). Recognising the difficulty of delivering practical training in an online environment (Microsoft Teams and Zoom) during the online events, several members of the team were present which enabled participants to go to break-out rooms with an expert if they needed additional help or wanted to discuss things. Throughout the project we sought feedback from the participants and used it to improve the tutorials prior to the following events. The feedback was obtained through two surveys using Qualtrics (\url{https://www.qualtrics.com/}). The participants of the events were asked to respond to one survey before the start of the event and another one after the end of the event. The first survey focused on the background of the participants and their original level of knowledge in relation to ABM (see appendix 1 for the questions). At the end of the event the participants were asked to answer a second survey. Questions of this survey were aimed at measuring the effectiveness of the tutorials and getting feedback on the workshop and tutorials (see appendix 2 for the questions). For some events participants had to register beforehand and we were able to send the pre-workshop survey by email. At other events no registration was possible or necessary and the pre-workshop survey was shared at the start of the event. The post-workshop survey was distributed using QR-codes or links at the end of the event.

@@

The surveys were analysed using R (R Core Team 2023). The following packages were used for analyses: ggplot2 (Wickham 2016), dplyr (Wickham et al. 2023), tidyr (Wickham, Vaughan and Girlich 2023), forcats (Wickham 2023a), lubridate (Grolemund and Wickham 2011) and stringr (Wickham 2023b). For the sake of reproducibility (Marwick 2017) this paper was written in R Markdown (\url{https://rmarkdown.rstudio.com/}), with data and code available at \url{https://github.com/ABMArchaeologists/ABMA_paper} (Visser et al. 2024b).

Over the course of the project three groups of students worked on the project during the Smart Solutions Semester at Saxion University of Applied Sciences. This is an interdisciplinary semester in which students of at least three different study-programmes or disciplines work together on a complex problem/project (\url{https://www.saxion.edu/business-and-research/collaborate-with-saxion/smart-solutions}). Within the semester students have to come up with new ideas or solutions to help a client with a wicked, or at least challenging, problem. The backgrounds of the various students we worked with were diverse: Applied Computer Science, Archaeology, Business Management Studies, Creative Business, Creative Media \& Game Technologies, and ICT. The aim of involving interdisciplinary groups in the project was to involve students in current research projects, to enable them to learn new things and share their innovative ideas and solutions with us. The outcomes of the students groups were supplementary to the deliverables of the project, but their new ideas were beneficial to the project. Their help during workshops was useful, but they also were engaged in developing educational material (Aalpoel et al. 2024), testing the tutorials, and developing a style for the website (Jutte et al. 2024) and other materials.

@@

The ABMA tutorials are publicly hosted on GitHub (\url{https://github.com/ABMArchaeologists/ABMA_tutorials/}) and are citeable (Rocks-Macqueen et al. 2024). Anyone interested can download the tutorials and their source code. To increase the accessibility and visibility of the tutorials, they are also hosted on a static website (\url{https://abmarchaeologists.github.io/ABMA_website}) and can be followed without the need to install any software. The source code of the website is also openly available (Jutte et al. 2024).

Besides the tutorial, the website provides: a link that can be used for feedback, basic explanations of agent-based modelling, an introductory presentation from one of the workshops, videos from an online course at Leiden University (Scherjon, Romanowska and Lambers 2019) and external resources that people can use to learn agent-based modelling. The website was mainly developed by a student group from Saxion University of Applied Sciences (Jutte et al. 2024).

@@

The final set of ABMA teaching materials comprises, next to accompanying materials (see above), the following tutorials (Rocks-Macqueen et al. 2024):

@@

In tutorial 2 the user learns the basics of NetLogo by building their first own simulation modelled on the Out of Africal dispersal of \emph{Homo sapiens} (Young and Bettinger 1995). The learner works through the basics of NetLogo syntax and learns how to set up a simulation and visualize its outcomes. This tutorial consists of nine lessons that guide the learner from the introduction to the beginner level of proficiency. The learning curve is relatively gentle. This is achieved by working with the NetLogo web interface that enables highlighting specific words and buttons and giving visual cues as to how to proceed. The tutorial introduces several commonly used built-in functions in Netlogo (called \emph{primitives}) and introduces the general structure of all models consisting of the initialization phase and the main simulation loop. It walks the participants through the most important features of the NetLogo world, such as dimensions, coordinates and the origin point of the grid on which the simulation runs. The learners gain their first stripes in programming by using simple loops and functions as well as conditional statements. The next level is achieved by introducing the topic of function definition and different types of variables. Exporting data produced by the model through plots is also explained.

In the third tutorial, the learner builds a simple trade model, again based on a published simulation (Romanowska 2018). Here learners expand their NetLogo programming skills with more complex syntactic structures such as loops, lists and reporters. They are introduced to some key computational techniques like modular code development and debugging which become important alongside an increased coding complexity. This tutorial consists of seven lessons. It intends to guide the learner from the beginner to the intermediate level of proficiency. This is achieved by introducing standards of code development such as modular code, pseudocode, debugging and annotating the code. In addition, the tutorial introduces custom agent breeds, visualization with labels and reporters, plots and monitors. This advances the proficiency level of the learner, since it develops knowledge of the programming language combined with diagnosing unexpected behaviour and problem solving.

@@

A large proportion of the participants of the workshops gave us information using the survey before the workshops (172 of 368 participants). The respondents came from at least 40 countries that represented almost all continents (see Figure \ref{fig:nationality}). There was a clear skew towards participants from Europe. The number of female respondents slightly outnumbered the male ones and a small group did not share their gender, while two identified as non-binary (see Figure \ref{fig:gender-age}). The ages of the respondents ranged from below 20 to over 70, although the majority fell between 20 and 40 years of age, indicating that the audience was predominantly composed of early career archaeologists. It seems that the different events also had a slightly different distribution of both gender and age. For example, more older people attended the workshop at the CAA conference in April 2023 and more men were present during the workshop at the Reuvensdagen in November 2023. This might be due to the differences of audiences at the conferences. For the Reuvensdagen the number of participants of the workshop was relatively low and this variance might be due to chance.

@@

Most of the respondents reported having a level of computer literacy already including particular computational skills common in archaeology, for example working with word processors, GIS and spreadsheets (see Figure \ref{fig:computer-skills}), and 111 of the respondents had some prior knowledge of ABM, while 59 were completely new to the subject. The majority (158) had never applied ABM to their research before participating in the workshops with only 14 participants who had experience of developing archaeological ABM. It is interesting to note that many respondents (75) could mention at least one type of software to be used for ABM (see Figure \ref{fig:abm-knowledge}).

@@

As shown above, the respondents had some knowledge on ABM in general, but did not know how to apply it or had never applied it before. The respondents were also asked how they rated the available theory on ABM (Figure \ref{fig:available-theory}). While 64 of them had no opinion on the subject, a proportion of the respondents (28) answered that they rated the available theory on ABM as limited and only a small group (39) as sufficient or better. This clearly shows the need for more and better educational material.

@@

\hypertarget{during-the-workshops}{%
\subsubsection{During the workshops}\label{during-the-workshops}}

During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials (Figure \ref{fig:workshop-impression}). Some participants engaged in discussion with the teachers to learn how they could apply ABM in their research or they discussed possibilities for the application of ABM in general. Individual progress varied depending on skills and preferences of the participants. While bugs were still a problem in earlier workshops, participant-teacher interaction was more focused on content-related learning problems in later installments.

@@

A similar proportion of the participants of the workshops gave us feedback after the workshops (171 of 368 participants).

@@

Most respondents (153) were enthusiastic about the teaching materials, with the majority rating them as excellent or good. The teachers were rated even better than the teaching materials, with most respondents (165) giving them the \emph{excellent} score (see Figure \ref{fig:rating-teaching}).

@@

The respondents generally liked the interactive step-by-step way of going through the tutorials, but also valued the interaction with the teachers strongly. The easy and intuitive introduction to ABM and NetLogo were also mentioned quite often. This was also observed during the workshops as most participants were working at their own pace and did not need much help. From a didactic point of view, the self-guided and self-paced way of going through the tutorials is very efficient for teachers and limits the frustration of students that may otherwise go too fast or fall behind the class when the tempo is dictated by the teacher.

@@

Beyond the creation of the OERs, the project also worked on aligning the learning outcomes of the OERs with the Digital Competence Framework for Citizens and the Digital Archaeological Skills Passport.

@@

As mentioned above, the following competence areas were identified as relevant for the tutorials: \emph{1. Information and data literacy}, \emph{2. Communication and collaboration}, \emph{3. Digital content creation} and, \emph{5. Problem solving} (European Commission, Joint Research Centre et al. 2022). In these competence areas various competences are developed towards an advanced or even specialised level. A more extensive description of the Digital Competence Framework for Citizens in relation to Agent-Based Modelling for archaeologists is available elsewhere (Visser et al. 2024a).

For \emph{Competence area 1. Information and data literacy}, the tutorials mainly address \emph{1.2 Evaluating Data, Information And Digital Content} and \emph{1.3 Managing Data, Information And Digital Content}. In the ABMA educational materials, users learn about working with data, being critical of their digital content and how to model and manage the data in the context of an ABM. For tutorial 2 the user is expected to be able to find information using the NetLogo(Web) dictionary (\url{http://ccl.northwestern.edu/netlogo/bind/} or \url{http://ccl.northwestern.edu/netlogo/docs/dictionary.html}).

For \emph{Competence Area 2. Communication and collaboration} the tutorials mainly address \emph{2.1 Interacting Through Digital Technologies}, because the users are constantly interacting with digital technologies, but communication is not relevant for the tutorials. However, during the workshops given in the course of the project, participants interacted with each other and the teachers.
